### 2024-06-19 README.Rmd
First full draft completed

```{r include=FALSE}
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(knitr)
library(magick) # Advanced Graphics and Image-Processing in R
library(neuralnet) # Training of Neural Networks
```

@@

By: Antonio Páez (paezha@mcmaster.ca)

@@

@@

```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
knitr::include_graphics("images/ai-bug-fix.png")
```

@@

```{r ai-litter-box, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 2. Revolutionizing pet health"}
knitr::include_graphics("images/ai-litter-box.png")
```

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 80 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, consuming fully one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

@@

```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 5. Simple Example of a Shallow Neural Network"}
knitr::include_graphics("images/nn1.png")
```

@@

```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 7. Example of a Deeper Neural Network"}
knitr::include_graphics("images/nn2.png")
```

@@

```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 9. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-econometrics.png")
```

```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 10. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-cell-biology.png")
```

```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 11. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
knitr::include_graphics("images/mac-library-euphemism.png")
```

@@

This is an utterly unrealistic expectation (also see automated grading below). Ideally, students should have the opportunity to discuss their drafted work with instructors before final submission (or again, could ask members of their study groups to proof-read their work, among many other ways to get feedback before submission). A conscientious instructor will invest time reading and understanding the draft work, so as to be able to provide quality feedback. But there is a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case? Whose data are used to train the machine that provides the feedback? Who fine-tunes the machine that provides the feedback? Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes. A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.]?

@@

```{r flickr-potato, out.width="50%", fig.cap="Figure 12. Data analysis using a chatbot? Maybe not such a good idea"}
knitr::include_graphics("images/potato.png")
```

@@

```{r puzzle-web-search, eval=FALSE, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
ggdraw() + 
  draw_image(image = "images/logical-puzzle-web-search.png")
```

@@

ggdraw() + 
  draw_plot(puzzle_plot) +
  draw_image(image = "images/logical-puzzle-chat-gpt-4.png")

@@

```{r google-reddit, out.width="50%", fig.cap="Figure 13. Google's AI Overview result was traced back to a post by redditor Fucksmith"}
knitr::include_graphics("images/google-cheese-glue.png")
```

@@

```{r google-the-onion, out.width="50%", fig.cap="Figure 14. Google's AI Overview result was traced back to an article in The Onion"}
knitr::include_graphics("images/google-the-rock.jpg")
```

@@

The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs. We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to duller versions of their own previous outputs. And this is before we even consider the possibility of AIs being deliberately "poisoned". As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023]. One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts. Another tool is @xu_shadowcast_2024's _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.

@@

@@

ChatGPT 4.o was announced on May 13, 2024, with fanfare and a flashy display of the model's new capabilities, which included the ability to respond vocally to prompts, and to identify images via a mobile device's camera. 4.o's voice, in particular, caused some titillation in the nether parts of even its creators [@roose_is_2024]. The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson. This film (to approximately no one's surprise) is, by his own account, one of OpenAI's Altman's favorites. The background story of 4.o's voice is somewhat sordid: Altman reportedly tried to hire Ms. Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when Ms. Johannson declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways. This voice was retired (probably to the profound sadness of many an AI engineer) after Ms. Johansson threatened legal action^[Altman's denials that OpenAI had not used Ms. Johansson's voice sounded hollow, after he [posted](https://x.com/sama/status/1790075827666796666) on social media the word "her" at the time the new version of ChatGPT was announced.] [@allyn_scarlett_2024].

Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI. Most people in the world lack that power. In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )

@germain_google_2023
@norem_intels_2023

@@

@perrigo_2_2023 (Follows summary)

ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified. Trained on horrible content, it was hard to sell ChatGPT when it was prone to racist, sexist, obscene outbursts. But companies like facebook, who did not want to pay for moderation, had hit on a solution of sorts <!--: train an AI to detect content that was unpalatable.-->

The solution?

> [F]eed an AI with labeled examples of violence, hate speech, and sexual abuse, and that tool could learn to detect those forms of toxicity in the wild. That detector would be built into ChatGPT to check whether it was echoing the toxicity of its training data, and filter it out before it ever reached the user. It could also help scrub toxic text from the training datasets of future AI models.

The key here is "labeled examples". What are those labels? Many machine algorithms do not learn by themselves...they learn from humans, and need what is called "labelled" data: a images that humans have labelled as "dog", "cats", etc. Or bits of text that have been labelled as...well...questionable. Labelling data is extremely time consuming and labor intensive. What's a corporation to do?

> To get those labels, OpenAI sent tens of thousands of snippets of text to an outsourcing firm in Kenya, beginning in November 2021. Much of that text appeared to have been pulled from the darkest recesses of the internet. Some of it described situations in graphic detail like child sexual abuse, bestiality, murder, suicide, torture, self harm, and incest. "

What did this mean to the people doing all this labeling in one of the poorest regions in the world? First of all, an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?

> One Sama worker tasked with reading and labeling text for OpenAI told TIME he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child. “That was torture,” he said. “You will read a number of statements like that all through the week. By the time it gets to Friday, you are disturbed from thinking through that picture."

@@

> Karen Hao: You can interact with ChatGPT in Swahili, but ChatGPT was developed primarily to work in English. So a lot of the scrubbing, the content moderation, the important safety measures within the chatbot were done in English. So when you prompt it in Swahili, you'll get more misinformation. You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.

@roberts_commercial_2016 Commercial {Content} {Moderation}: {Digital} {Laborers}' {Dirty} {Work}

@@

@@

Where will new data come from? [@murgia_why_2023]

Note how several organizations have responded to the AIfication of all the things: by banning it. The concerns cited are seldom ethical or moral, but mostly about the risk of sensitive information leaking. As LLMs begin to be increasingly trained on model-produce data, overfitting (or memorization) can become a larger risk.

And this is just a result of using an overfitted model. What about data leaks? Some stories regarding Microsoft's security track record:

https://www.theregister.com/2024/04/12/microsoft_cisa_order/

https://techreport.com/news/microsoft-azure-hit-with-the-largest-data-breach-in-its-history-hundreds-of-executive-accounts-compromised/

Also, how confident can the public be about assurances that a big tech corporation will not data mine and commercialize data by the companies that sell the services of LLMs? Microsoft is so big that it can stand up to the US government:

https://www.theregister.com/2024/04/05/microsoft_government_contracts/

<!-- "Nice operation you have there...be a shame if your systems stopped working all of a sudden."-->

Organization ranging from Free Open Source projects to elements of the federal government in the United States have limited/restricted the use of AI.

NetBSD (a project that produces a free, Unix-like Open Source operating system) states in its commit guidelines (contributions to code) that "[c]ode generated by a large language model or similar technology...is presumed to be tainted code, and must not be committed without prior written approval by core." [@netbsd_netbsd_nodate].

The National Archives and Records Administration (USA) bans use of ChatGPT [@koebler_national_2024]

US Congress bans use of ChatGPT and Copilot for staff [@singh_after_2024]. This article also note how companies like Samsung and Apple have restricted the use of chatbots citing concerns with data security.

@@

I struggle a bit with this one. For example, I fly more often than the [typical Canadian](https://ourworldindata.org/grapher/air-trips-per-capita?tab=table&time=earliest..latest&country=~CAN) and a little bit less often than the [typical world traveller](https://www.news24.com/news24/around-the-world-in-50-hours-average-person-takes-65-flights-per-year-20180828) (but I have not owned a car in decades). Who am I to dictate how much damage _other people_ have a right to inflict (or not) to the environment.

Microsoft: Here's how burning the planet will help to save it.

https://blogs.microsoft.com/on-the-issues/2024/05/15/microsoft-environmental-sustainability-report-2024/ [@smith_our_2024]

https://blogs.microsoft.com/on-the-issues/2023/11/16/accelerating-sustainability-ai-playbook/ [@nakagawa_accelerating_2023]

https://www.theguardian.com/commentisfree/article/2024/may/30/ugly-truth-ai-chatgpt-guzzling-resources-environment

"Environmental costs of AI are soaring" [@crawford_generative_2024]

@@

ChatGPT consumes substantially more energy than a regular web search, up to 25 times more, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024].

@@

The example above may seem like an amusement, but a GenAI is not a toy, as Air Canada discover when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].

@@

But what when an AI replicates the both-sides perspective of much of the media when trying to learn about an existential crisis like climate changes? [@gordon_multiple_2024]. Or when trying to obtain accurate voting information in the context of voter supression? [@angwin_seeking_2024].

@@

But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral. Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017]. Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020]. A U.S. Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S. politics [@us_department_of_state_gec_2024, p. 51].]. Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021. In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].]. The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].

Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023]. In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda. Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].]. In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr. Obama [@tenbarge_glue_2024]. Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024]. This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do). Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring. Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China. These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines. To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI. VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024]. Who should be concerned about maintaining an edge?]?

```{r google-obama, out.width="50%", fig.cap="\\label{fig:google-ai-overview-of-obama}Figure 17. Google's AI Overview of Muslim US presidents"}
google_plot <- ggplot() +
  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
  theme_void()

ggdraw() + 
  draw_plot(google_plot) +
  draw_image(image = "images/google-obama.png")
```

@@

AI industry and the military: @haskins_im_2024

## Bonus

```{r googly-eyes, out.width="50%", fig.cap="\\label{fig:googly-ai-eyes}Figure 18. In conclusion"}
google_plot <- ggplot() +
  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
  theme_void()

>>>>

```{r include=FALSE}
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(glue)
library(here)
library(knitr)
library(magick) # Advanced Graphics and Image-Processing in R
library(neuralnet) # Training of Neural Networks
```

@@

Antonio Páez (paezha@mcmaster.ca)

@@

Note: _This essay was written as an open research project, and it can be found in the following repository:_ https://github.com/paezha/University-AI-Panel

@@

```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
knitr::include_graphics(glue::glue(here::here(), "/images/ai-bug-fix.png"))
```

@@

```{r ai-litter-box, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 2. Revolutionizing pet health"}
knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
```

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 90 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, consuming fully one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

@@

```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 5. Simple Example of a Shallow Neural Network"}
knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
```

@@

```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 7. Example of a Deeper Neural Network"}
knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
```

@@

```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 9. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-econometrics.png"))
```

```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 10. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-cell-biology.png"))
```

```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 11. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-euphemism.png"))
```

@@

This is an utterly unrealistic expectation (also see automated grading below). Ideally, students should have the opportunity to discuss their drafted work with instructors before final submission (or again, could ask members of their study groups to proof-read their work, among many other ways to get feedback before submission). A conscientious instructor will invest time reading and understanding the draft work, so as to be able to provide quality feedback. But there is a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case? Whose data are used to train the machine that provides the feedback? Who fine-tunes the machine that provides the feedback? Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes^[GenAI is not a toy, as Air Canada discovered when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot it deployed was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].]. A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.]?

@@

```{r flickr-potato, out.width="50%", fig.cap="Figure 12. Data analysis using a chatbot? Maybe not such a good idea"}
knitr::include_graphics(glue::glue(here::here(), "/images/potato.png"))
```

@@

```{r puzzle-web-search, eval=FALSE, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
ggdraw() + 
  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-web-search.png"))
```

@@

ggdraw() + 
  draw_plot(puzzle_plot) +
  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-chat-gpt-4.png"))

@@

```{r google-reddit, out.width="50%", fig.cap="Figure 13. Google's AI Overview result was traced back to a post by redditor Fucksmith"}
knitr::include_graphics(glue::glue(here::here(), "/images/google-cheese-glue.png"))
```

@@

```{r google-the-onion, out.width="50%", fig.cap="Figure 14. Google's AI Overview result was traced back to an article in The Onion"}
knitr::include_graphics(glue::glue(here::here(), "/images/google-the-rock.jpg"))
```

@@

The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs. We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to duller versions of their own previous outputs. And this is before we even consider the possibility of AIs being deliberately "poisoned". As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023]. One such tool is _PhotoGuard_, developed by a team at MIT to send signals to an AI that prevents it from manipulating images [@heikkila_this_2023]. Other examples include @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts. And more recently, @xu_shadowcast_2024 presented _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.

@@

## What do GenAIs do for the Province?

As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.

But AIs must do something for someone, otherwise why would they be everywhere? Given history, one could be forgiven to think that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education. GenAIs appear on the horizon as an  innovation-sounding rationale (a mandate even!) for the province to force institutions of higher education to find "efficiencies". Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of the technology's flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023]. Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], one would hav to be blind not to see the glamour that GenAI will cast on governments that do not seem to believe in higher education in the first place.

```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}'I like the look of his cuts'"}
google_plot <- ggplot() +
  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
  theme_void()  +
  theme(text = element_text(size = 8))

ggdraw() + 
  draw_plot(google_plot) +
  draw_image(image = glue::glue(here::here(), "/images/ai-googly-eyes.png", width = 0.8))
```

@@

ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities. These included the chatbot's ability to respond vocally to prompts, and to identify images via a mobile device's camera. 4.o's voice, in particular, caused some titillation in the nether parts of even its creators [@roose_is_2024]. The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson. By Sam Altman's own account, this is one of his favorite movies. The background story of 4.o's voice is somewhat sordid: Altman reportedly tried to hire Ms. Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when she declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways. This voice was retired (probably to the profound sadness of many an AI engineer) after Ms. Johansson threatened legal action^[Altman's denials that OpenAI had not used Ms. Johansson's voice sounded hollow, after he [posted](https://x.com/sama/status/1790075827666796666) on social media the word "her" at the time the new version of ChatGPT was announced.] [@allyn_scarlett_2024].

Alas, very few have the social standing and wealth of Scarlett Johansson to challenge OpenAI. Most people in the world lack that power and the AI industry has not been shy to exploit that imbalance [@roberts_commercial_2016], with Google's privacy policy for using the whole internet as the source of data for its AI efforts [@germain_google_2023], and Adobe landing in hot water by adopting terms of use widely thought to be confiscatory [@weatherbed_adobes_2024].

Besides being extractive, the AI industry is exploitative in darker ways. ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified. Trained on horrible content, it was hard to sell ChatGPT when it was prone to racist, sexist, obscene outbursts. Fortunately (or unfortunately, depending on perspective) companies like facebook, who did not want to pay for moderation, had already hit on a solution of sorts, which was basically to train AIs to detect content that was unpalatable. Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity in the wild, to filter it out before it reached users. However, to obtain the essential labeled inputs, OpenAI sent tens of thousands of samples to Kenya, with text that described murder, suicide, bestiality, self-harm, incest, torture, and child sexual abuse situations in graphic detail.

What did this mean to the people doing all the labeling in one of the poorest regions in the world? First of all, an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost? As one worker tasked with reading and labeling text for OpenAI told TIME, "he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child. 'That was torture,' he said. 'You will read a number of statements like that all through the week. By the time it gets to Friday, you are disturbed from thinking through that picture'" [@perrigo_2_2023].

<!--

In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )

@norem_intels_2023

@@

@@

> Karen Hao: You can interact with ChatGPT in Swahili, but ChatGPT was developed primarily to work in English. So a lot of the scrubbing, the content moderation, the important safety measures within the chatbot were done in English. So when you prompt it in Swahili, you'll get more misinformation. You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.
-->

@@

### Privacy issues

@@

This means that data cannot be presumed to be private once that it has been used to train an AI. In response organizations such as the National Archives and Records Administration in the US [@koebler_national_2024], and the US Congress [@singh_after_2024] have banned the use of ChatGPT and Copilot. @singh_after_2024 also notes how companies like Samsung and Apple have restricted their use of chatbots.

NetBSD (a project that produces a free, Unix-like Open Source operating system) states in its commit guidelines (contributions to code) that "[c]ode generated by a large language model or similar technology...is presumed to be tainted code, and must not be committed without prior written approval by core." [@netbsd_netbsd_nodate]. The concerns cited are seldom ethical or moral, but mostly about the risk of sensitive information leaking. As LLMs begin to be trained on model-produced data [@murgia_why_2023], overfitting (or memorization) will likely become a larger risk.

And this is just a result of using an overfitted model. What about data leaks? Microsoft's data security [track](https://www.theregister.com/2024/04/12/microsoft_cisa_order/) [record](https://techreport.com/news/microsoft-azure-hit-with-the-largest-data-breach-in-its-history-hundreds-of-executive-accounts-compromised/) does not inspire confidence.

@@

<!--
I struggle a bit with this one. For example, I fly more often than the [typical Canadian](https://ourworldindata.org/grapher/air-trips-per-capita?tab=table&time=earliest..latest&country=~CAN) and a little bit less often than the [typical world traveller](https://www.news24.com/news24/around-the-world-in-50-hours-average-person-takes-65-flights-per-year-20180828) (but I have not owned a car in decades). Who am I to dictate how much damage _other people_ have a right to inflict (or not) to the environment.
-->

Every human activity has an environmental impact, and at a time when we face unprecedented challenges with climate change, we must be discerning in how we consume resources. The choice is so clear that Microsoft has basically told us: we need to burn the planet [@smith_our_2024] to create the AIs that will help us save the planet [@nakagawa_accelerating_2023].

@crawford_generative_2024 has already sounded the alarm: the environmental costs of AI are soaring, and doing so mostly in secret [@luccioni_power_2024]. It has been estimated that ChatGPT consumes substantially more energy than a regular web search, up to 25 times more, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024]. As the examples in the preceding sections illustrate, this is in exchange for results that are not better than a simple search, and often are worse. Widespread AI deployment may in fact make the climate crisis worse, not only by guzzling resources, but also by spreading misinformation [@gordon_multiple_2024].

<!--
https://www.theguardian.com/commentisfree/article/2024/may/30/ugly-truth-ai-chatgpt-guzzling-resources-environment

@@

@@

The example above may seem like an amusement, but a .

@@

But what when an AI replicates the both-sides perspective of much of the media when trying to learn about an existential crisis like climate changes? [@gordon_multiple_2024]. Or when trying to obtain accurate voting information in the context of voter supression? [@angwin_seeking_2024].
-->

@@

But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral. Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017]. Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020]. A U.S. Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S. politics [@us_department_of_state_gec_2024, p. 51].]. Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021. In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].]. The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021]. This is not a bug of the system, but rather a feature, as Meta's supression of information about Gaza in 2024 shows [@paul_former_2024].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].

Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023]. In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda. Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].]. In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr. Obama [@tenbarge_glue_2024]. Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024]. This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do). Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring. Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China. These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines. To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI. VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024]. Who is concerned about maintaining an edge?]?

```{r google-obama, out.width="50%", fig.cap="\\label{fig:google-ai-overview-of-obama}Figure 17. Google's AI Overview of Muslim US presidents"}
google_plot <- ggplot() +
  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
  theme_void() +
  theme(text = element_text(size = 8))

ggdraw() + 
  draw_plot(google_plot) +
  draw_image(image = glue::glue(here::here(), "/images/google-obama.png"))
```

@@

AI industry and the military: @haskins_im_2024
-->

## Appendices
